Collaborative Visual Analysis on the Web with RCloud

### 2015-06-26 conclusion.tex
chaff

RCloud was readily accepted as a platform for deployment of
visualizations and interactive exploration tools inside our organization.  
Notebook sharing and publishing were eagerly adopted.
Features for single-user data exploration, on the other hand, that
compete with existing mature tools, were not readily accepted.
In the future we hope to study and understand the degree to which
this reluctance is caused by bad design decisions on our part, and how
much comes from ``mere'' change aversion. It is also clear we need
need better ways to manage the persistence of notebooks and to cope
with potential information overload.

@@

RCloud is open-source, and available at \url{github.com/att/rcloud/}.

>>>>

RCloud was readily accepted as a platform for deployment of
visualizations and interactive exploration tools inside our organization.  
Notebook sharing and publishing were eagerly adopted.
On the other hand, features for single-user data exploration that
compete with existing mature tools, were not readily accepted.
In the future we hope to study and understand the degree to which
this reluctance is caused by bad design decisions on our part, and how
much comes from ``mere'' change aversion. It is also clear we need
need better ways to manage the persistence of notebooks and to cope
with potential information overload.

@@

RCloud is available at \url{github.com/att/rcloud/}.

### 2015-06-26 discussion.tex
remove comments
chaff

\begin{table}
  \centering
  \begin{tabular}{l|cccccc}
    & V/F & C & D & ML & R & A \\
    \hline
RCloud     &       x            &       x       &     x      &      x        &           x        &           x         \\
RStudio    &                    &               &            &               &           x        &           x         \\
JSFiddle   &       x            &       x       &            &               &                    &                     \\
bl.ocks    &       x            &       x       &            &               &           x        &                     \\
shiny      &                    &               &     x      &               &           x        &           x         \\
Jupyter    &                    &               &            &      x        &           x        &           x         \\
Tableau    &                    &      x        &     x      &               &           x        &                     \\
ManyEyes   &                    &      x        &     x      &               &                    &                     
  \end{tabular}
  \caption{Comparison of system features. Columns: Versioning/Forking, Collaboration, Deployment, Multi-language support, Integrated Reports, and Integrated Analysis.}
\end{table}
% SN says Tableau: versioning=no  collaboration=yes  deployment=yes multilanguage=no  reports=yes  analysis=no

@@

\paragraph*{Versioning and Upgrading}
Some aspects of the environment are more difficult to
manage in RCloud than in conventional systems. RCloud does not separate
its development and deployment environments, and every version of every
prototype is shared by default. Although this encourages
collaboration, it also quickly exposes misconfigurations,
mismatches between package versions, and programming errors
that can unintentionally affect production websites.
If an RCloud website is distributed across multiple servers,
conflicting versions of system software
can also cause errors that are hard to debug.

Versioning of software components must be managed in several
levels: in data analysis and visualization scripts (the
notebooks themselves); in the R environment such as the installed
libraries and packages; and in the external environment such as the
operating system, libraries, and protocols spoken by remote
services. At one end of the scale, we have full control over notebooks
via git, along with conventions to fix stable notebook versions.
The R environment is under the control of its package manager that
has rules to ensure reasonable consistency.
% (It is possible for RCloud to access information about package
% versions and configurations, but support for checking compatibility
% and maintaining multiple versions in the same environment is not
% strong.)
At the other end of the scale, there is basically no
control over the external environment, and if an external service 
used by a notebook changes its API, the notebook can stop working.

@@

\paragraph*{Caching}
An important issue is how and where to introduce caching in
RCloud's distributed computation model. Caching can dramatically
improve performance in a way that is transparent to
application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
Currently, RCloud offers no support for implicit caching, but 
it's possible to incorporate caching in individual R libraries.
The reuse of intermediate results is a recurring problem
in many data exploration environments.
We envision general cache management being implemented in a
new middle layer in the future.

\paragraph*{Security}
Because RCloud allows arbitrary R code to be run
\emph{by design}, it's not straightforward to protect deployments
from running unauthorized arbitrary code. RCloud uses an object capability model
\cite{Miller:2006:RCT} recently added to the Rserve
protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
to remote procedure calls; web browsers never directly instruct the
RCloud backend to execute arbitrary code, preventing unauthenticated clients
from making unauthorized calls to the RCloud runtime environment.
%
Our back-end environment, on the other hand, assumes a high degree of trust between users:
access control is delegated to the host operating system and web server.
Most operating systems rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome.
Information security in RCloud today is an unsolved matter.
More sophisticated approaches are open research topics~\cite{Moore:2011:SAF}.
% We are hopeful that having a richer
% environment for collaboration and information sharing will encourage
% new approaches to information security in visual analytics.
%\gordon{This sounds like we think we are helping by pushing the
%system to the breaking point, is that the idea?}

@@

% Surprisingly, we found that when users realize that \emph{scratch work
We found that when users realize that \emph{scratch work
is getting saved}, its sheer volume becomes burdensome.
Clearly, RCloud is an extreme, and the right solution is
probably in the middle. An intriguing possibility,
suggested by one of our interviewees, is to implement auditing of data
analyses as it existed in S~\cite{Becker:1988:Auditing}, tying
data artifacts to the processes that generated them. 
Bookkeeping would be reduced to selecting the \emph{data} to be kept,
and the code that generated it would be derived and saved transparently.

%% . This is why we did not implement a simple command line
%% as many users request. We could imagine making an exception for commands
%% that have no side-effects, but in the general case this is no easier than
%% auditing because R is not a purely functional language.

%% The interface is also perceived to be slower, both for computer and for
%% human. First, there is higher latency when commands run, because
%% RCloud saves a cell to a repository before executing it. It
%% may be possible to safely write to the repository behind execution of the
%% same code. Second, and perhaps more detrimental, the interface requires
%% switching between the keyboard and mouse, which can be cumbersome.
%% %\gordon{we don't need to back that up, right? i hate using passive voice.}
%% Shortcut keys can defined as a workaround, but it would be better
%% to reduce the need for them.

%% Other complaints mostly focus on layout, and can probably be addressed by
%% offering better customization, and designs such as tabbed views.

\paragraph*{Discovery is hard} 
With hundreds of users and thousands of notebooks, a global notebook tree
becomes unwieldy. The search function does not protect searchers from
obsolete and erroneous notebooks, which by default are retained
forever. In addition, important metadata (has this notebook been
recently run, or edited?) is not easily searchable. In other words,
\emph{textual search} is easy, but \emph{relevance search} is what we
need. Allowing users to \emph{tag} notebooks is natural, but
relies on users making a conscious effort, which goes against
our ideal of not asking the user to do work for the system.
% It is better to have the system work for the user,
% relying on passive metrics, such as how often a notebook has been run.
% More generally, infrastructure for automatic
Automatic
\emph{curation} of
software artifacts generated by teams seems to be an important avenue
for future work.

>>>>

\begin{table}
  \centering
  \begin{tabular}{l|cccccc}
    & V/F & C & D & ML & R & A \\
    \hline
RCloud     &       x            &       x       &     x      &      x        &           x        &           x         \\
RStudio    &                    &               &            &               &           x        &           x         \\
JSFiddle   &       x            &       x       &            &               &                    &                     \\
bl.ocks    &       x            &       x       &            &               &           x        &                     \\
shiny      &                    &               &     x      &               &           x        &           x         \\
Jupyter    &                    &               &            &      x        &           x        &           x         \\
Tableau    &                    &      x        &     x      &               &           x        &                     \\
ManyEyes   &                    &      x        &     x      &               &                    &                     
  \end{tabular}
  \caption{Comparison of system features. Columns: Versioning/Forking, Collaboration, Deployment, Multi-language support, Integrated Reports, and Integrated Analysis.}
\end{table}

@@

\paragraph*{Versioning and Upgrading}
Some aspects of the environment are more difficult to
manage in RCloud than in conventional systems. RCloud does not separate
its development and deployment environments, and every version of every
notebook is shared by default. Although this encourages
collaboration, it also quickly exposes misconfigurations,
mismatches between package versions, and programming errors
that can unintentionally affect production websites.
%% If an RCloud website is distributed across multiple servers,
%% conflicting versions of system software
%% can also cause errors that are hard to debug.
Versioning of software components must be managed in several levels: in
the notebooks themselves, in the installed libraries and packages of the
R environment, and in the external environment such as the operating
system, libraries, and protocols spoken by remote services.
%% At one end of the scale, we have full control over notebooks
%% via git, along with conventions to fix stable notebook versions.
%% The R environment is under the control of its package manager that
%% has rules to ensure reasonable consistency.
%% % (It is possible for RCloud to access information about package
%% % versions and configurations, but support for checking compatibility
%% % and maintaining multiple versions in the same environment is not
%% % strong.)
%% At the other end of the scale, there is basically no
%% control over the external environment, and if an external service 
%% used by a notebook changes its API, the notebook can stop working.

@@

\paragraph*{Security}
Because RCloud allows arbitrary R code to be run
\emph{by design}, it's not straightforward to protect deployments
from running unauthorized arbitrary code. RCloud uses an object capability model
\cite{Miller:2006:RCT} recently added to the Rserve
protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
to remote procedure calls: web browsers never directly instruct the
RCloud backend to execute arbitrary code, preventing unauthenticated clients
from making unauthorized calls to the RCloud runtime environment.
Our back-end environment, on the other hand, assumes a high degree of trust between users:
access control is delegated to the host operating system and web server.
Most operating systems rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome.
Information security in RCloud today is an unsolved matter.
More sophisticated approaches are open research topics~\cite{Moore:2011:SAF}.

@@

We found that when users realize that \emph{scratch work
is getting saved}, its sheer volume becomes burdensome.
Clearly, RCloud is an extreme, and the right solution is
probably in the middle. An intriguing possibility,
suggested by one of our interviewees, is to implement auditing of data
analyses as it existed in S~\cite{Becker:1988:Auditing}, tying
data artifacts to the processes that generated them. 
Bookkeeping would be reduced to selecting the \emph{data} to be kept,
and the code that generated it would be derived and saved transparently.

\paragraph*{Discovery is hard} 
With hundreds of users and thousands of notebooks, a global notebook tree
becomes unwieldy. The search function does not protect searchers from
obsolete and erroneous notebooks, which by default are retained
forever. In addition, important metadata (has this notebook been
recently run, or edited?) is not easily searchable. In other words,
\emph{textual search} is easy, but \emph{relevance search} is what we
need. Allowing users to \emph{tag} notebooks is natural, but
relies on users making a conscious effort, which goes against
our ideal of not asking the user to do work for the system.
Automatic
\emph{curation} of
software artifacts generated by teams seems to be an important avenue
for future work.
